VFTS682: a confirmed dynamical ejection?

### 2018-05-10 VFTS682.tex
new sec. origin

Based on the extremely high mass of this star and
its present-day apparent isolation, \cite{bestenlehner:11} proposed it
could be a candidate for isolated star formation, or a ``slow runaway'' ejected
from R136 in the past. Their study raised the interest in the origin
of this star. \cite{fujii:11, banerjee:12} carried out N-body
simulations of massive young stellar clusters and concluded that the
ejection of such a massive object from R136 through dynamical
interactions \citep[e,g,][]{poveda:67} would be possible.
The competing ejection mechanism, the disruption of a binary
system by a core-collapse event \citep[][]{zwicky:57, blaauw:61}, is
thought to be ruled out, since the cluster has an estimated age of
$\lesssim2$\,Myr \citep[][]{sabbi:12}, which is shorter than the shortest stellar lifetime
\citep[$\sim$3\,Myr, e.g.,][]{zapartas:17}.

In this study, we combine the radial velocity measurements from the
VFTS survey \citep[][]{evans:11} with the proper motion from Gaia DR2
to reconstruct the three-dimensional velocity of VFTS682, and test the
hypothesis that this star was ejected from R136. We discuss in
\Secref{sec:sample} the data for VFTS682, and the selection of stars
used to define a local reference frame. Our results
indicate that R136 is a bona fide runaway star (\Secref{sec:runaway}),
therefore isolated star formation is \emph{not}
required to explain it. We also find that a dynamical ejection from
R136 is compatible with the direction of its velocity
vector. 
We conclude
with a very brief discussion on the implications for theories of
star formation, N-body interactions, and binary evolution in
\Secref{sec:discussion}.

@@

Using the 651 stars selected as described in \Secref{sec:sample} and shown in blue in \Figref{fig:main}, we find averaged proper motion components of
$\langle\mu_\mathrm{RA}^\mathrm{sur}\rangle = 1.683\pm0.002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DE}^\mathrm{sur}\rangle =
0.672\pm0.003\,\mathrm{mas\ yr^{-1}}$. We note that these values are
in good agreement with what found by \cite{lennon:18}. Subtracting these values from the
proper motions of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu^\mathrm{sur}_\mathrm{RA} = 0.16\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu^\mathrm{sur}_\mathrm{DE} =
0.11\pm 0.08\,\mathrm{mas\ yr^{-1}}$. We note that the error budget is
dominated by the errors on the proper motion components of VFTS682.

@@

The kinematic age $\tau_\mathrm{kin}$ is smaller than the apparent age
of the star age $1.0\pm 0.2$\,Myr from \cite{schneider:18}, which
corroborates the idea that the star is the result of a dynamical
ejection. The value of $\tau_\mathrm{kin}$ we find is also smaller
than the corresponding value for another massive runaway in 30 Dor,
VFTS16 for which \cite{lennon:18} inferred a kinematic age of
$\sim$1.5\,Myr. This means that the more
massive VFTS682 was ejected later than VFTS16.

@@

\todo{compare to Lennon: time of ejection is the same? what does it
  say for cluster}
\todo{say implications for center of R136 (same stars are there), how
  many ejected given this one, was a binary ejected in the other direction?}
\todo{does this imply there is a binary in the core ?}
\begin{itemize}
\item VFTS682 is a bona fide runaway with $v\sim60\,\mathrm{km\
    s^{-1}}$ thrown out from R136. Both its speed and the age of the
  cluster are consitent with a dynamical ejection.
  \item VFTS682 comes from R136 as was expected by
  \cite{bestenlehner:11, fujii:11, banerjee:12}, so it does not
  require isolated SFH to be explained
\item apparent age tension (connect to VFTS16 as well).
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\end{itemize}
\section{Summary}

1. Gaia data confirm a 150Msun runaway with v~60, thus the most
massive known to date
2. directions suggest ejection from R136, confirming prediction of
fuji+ banerjee+
2b. This fits with the observation from crowther of M>300Msun in R136
(and mention vfts682 has the same spectral type): the most likely
mechanism needs encounter with 2 more massive -> additional evidence
for Crowther claim
3. comment on final fate (BH)

Random notes: $v\sin(i)<200\,\mathrm{km\ s^{-1}}$ form \cite{schneider:18}, age $1.0\pm 0.2$\,Myr
from \cite{schneider:18}

Many other very massive stars are present in the surroundings this cluster, and a more
detailed analysis on the larger sample is desirable.

>>>>

Based on the extremely high mass of this star and
its present-day apparent isolation, \cite{bestenlehner:11} proposed it
could be a candidate for isolated star formation, or a ``slow runaway'' ejected
from R136 in the past. Their study raised the interest in the origin
of this star. \cite{fujii:11, banerjee:12} carried out N-body
simulations of massive young stellar clusters and concluded that the
ejection of such a massive object from R136 through dynamical
interactions \citep[e,g,][]{poveda:67} would be possible.
The competing ejection mechanism, the disruption of a binary
system by a core-collapse event \citep[][]{zwicky:57, blaauw:61}, is
thought to be ruled out, since the cluster has an estimated age of
$\lesssim2$\,Myr \citep[][]{crowther:10,sabbi:12}, which is shorter than the shortest stellar lifetime
\citep[$\sim$3\,Myr, e.g.,][]{zapartas:17}.

In this study, we combine the radial velocity measurements from the
VFTS survey \citep[][]{evans:11} with the proper motion from Gaia DR2
to reconstruct the three-dimensional velocity of VFTS682, and test the
hypothesis that this star was ejected from R136. We discuss in
\Secref{sec:sample} the data for VFTS682, and the selection of stars
used to define a local reference frame. Our results
indicate that R136 is a bona fide runaway star (\Secref{sec:runaway}),
therefore isolated star formation is \emph{not}
required to explain it. We also find that a dynamical ejection from
R136 is compatible with the direction of its velocity
vector. 
We conclude
with a very brief discussion on the implications for star formation,
cluster evolution, and stellar physics studies in
\Secref{sec:discussion}.

@@

Using the 651 stars selected as described in \Secref{sec:sample} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}^\mathrm{sur}\rangle = 1.683\pm0.002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DE}^\mathrm{sur}\rangle =
0.672\pm0.003\,\mathrm{mas\ yr^{-1}}$. We note that these values are
in good agreement with what found by \cite{lennon:18}. Subtracting these values from the
proper motions of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu^\mathrm{sur}_\mathrm{RA} = 0.16\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu^\mathrm{sur}_\mathrm{DE} =
0.11\pm 0.08\,\mathrm{mas\ yr^{-1}}$. We note that the error budget is
dominated by the errors on the proper motion components of VFTS682.

@@

The kinematic age $\tau_\mathrm{kin}$ is smaller than the apparent age
of the star age $1.0\pm 0.2$\,Myr from \cite{schneider:18}, which
corroborates the idea that the star is the result of a dynamical
ejection.

@@

Based on our results, we can claim that VFTS682 is the most massive
runaway known to date, with a peculiar three-dimensional spatial
velocity of $\sim60\,\kms$. This means that isolated star formation is
\emph{not} required to explain this star. Its proper motion suggests that it was ejected from the cluster R136
$\sim0.6$\,Myr ago. Given the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

Dynamical ejections due to N-body interactions typically (although not necessarily) eject the least
massive star among those interacting \todo{ref}. This means that, just
based on the kinematic properties of VFTS682, we would expect several
stars with initial masses larger than $\sim150\,M_\odot$ in R136.
This can be considered an independent confirmation of the detection
of extremely massive stars by \cite{crowther:10} in the core of the
cluster. The projected rotational equatorial velocity of VFTS682
reported by \cite{schneider:18} is $v\sin(i)<200\,\kms$, which is in
line with the average rotation rate of massive stars in the region
\citep[][]{ramirez-agudelo:15}. This suggests that VFTS682 has not
experienced binary interactions, nor it will, since the multi-epoch
data of the VFTS survey rule out present-day binarity. Moreover, the spectral type of VFTS682
\citep[WNh5][]{bestenlehner:11} is the same as R136a1-a3, i.e.~the
three most massive stars detected in the core of the cluster by
\cite{crowther:10}. Therefore,
VFTS682 might be an ideal target to constrain the stellar physics of
stars with masses well above $\sim100\,M_\odot$: its isolation makes
it an easier target for observation than the crowded core of R136,
where similar stars exist.

The similarities between VFTS682 and the WNh5 stars in the core of
R136 are also in agreement with the ``bully binary'' model of
\cite{fujii:11}. They suggested, based on their numerical simulations,
that dynamical interactions lead to the formation of an extremely
massive binary in the early phases of a cluster evolution, and then it tightens by ejecting other stars passing
by. Interpreting our results for VFTS682 through the lens of their simulations would
suggest the presence of a binary with total mass
$M_1+M_2\gtrsim 400\,M_\odot$ in the core of the cluster, which they
suggested could be R145. The kinematic age of VFTS682 puts an
upperlimit to the timescale to form such ``bully binary'' in
R136. Unfortunately, their simulations did not include stars more
massive than $100\,M_\odot$, so it is difficult to predict, based on
the existence of VFTS682, how many other stars should have been
ejected from the cluster already and what is their mass
distribution. A comprehensive study of the kinematic properties of the
large sample of stars with extremely large inferred masses visible
around R136 is encouraged.

\cite{lennon:18} have carried out a similar study on VFTS16, another
massive star ($\sim90\,M_\odot$) in the 30 Doradus region, previously
known to be a runaway from the value of
its radial velocity. They also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster. 
The value of $\tau_\mathrm{kin}\simeq0.63$\,Myr we find for VFTS682 (see \Secref{sec:r136}) is smaller
than the corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic age of
$\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

\begin{itemize}
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\item comment on final fate (BH)

\end{itemize}
